Black swans in ecological time series\\ Black swans in population dynamics\\ Heavy-tailed ecological time series\\ Black swans in ecological dynamics\\ Black swan ecological dynamics\\ \ldots

### 2014-10-20 text/ms.tex
More intro tweaks

There are two key reasons why we may find little evidence of ecological black
swans. First, they might not exist. Indeed, the majority of population dynamics
model fitting and risk forecasting assumes that population dynamics are not
heavy tailed (e.g.~REF, REF, REF). Alternatively, dynamics might be heavy
tailed, but a paucity of statistical methods for the frequently available short
time series has limited our ability to detect them. For example, a commonly
applied probability distribution to detect the shape of time series tails (the
generalized extreme value distribution) requires lumping data into the most
extreme event per unit time (e.g.~heaviest rainfall per year) and therefore
requires a considerable quantity of data (REF). Previous large-scale work on
the distribution of abundance trends focused on model comparison to identify
parsimonious descriptions, but has not quantified the magnitude of heavy tails
on a continuous scale (REF). Furthermore, information criteria are poor at
distinguishing subtle differences in population dynamics for the relatively
short time series that are most readily available (REF CHANGE THIS?).
%bring in Ward - great method but data intensive, can't be applied to most
%individual time series from the GPDD
%plankton --- hideous distribution to fit, even harder in population dynamics
%framework -and what about seasonal cycles, density dependence?
%GPDD - model competition and finding most parsimonious explanation from set of
%hypotheses --- but only \emph{very} heavy from not heavy
%GEV --- super data intensive
%tipping point and regime shift methods --- say something\ldots{} not
%appropriate for most of the data available

Here we develop a probabilist framework for identifying ecological black swans
in population dynamic process noise --- the stochastic jumps from time step to
time step. Our framework quantifies both the magnitude and probability of
heavy-tailed dynamics, allows for a range of population dynamic models, can
incorporate observation uncertainty, and is applicable to commonly available
lengths of abundance time series. We apply our model to hundreds of populations
from around the world across four taxonomic classes to address three questions:
(1) how frequent and strong are black swans in population dynamics across
taxonomy, (2) what are some verified causes of black swans in ecological time
series, and (3) are there characteristics of time series or intrinsic
life-history characteristics associated with heavy-tailed dynamics. We find
a high probability of ecological black swans across a wide range of taxa but in
a limited number of populations. When they occurred, extreme climate,
predation, parasites, and their interactions were common documented causes. We
found little evidence that intrinsic life-history characteristics were related
to the probability of black swans.

>>>>

There are two key reasons why we may find little evidence of ecological black
swans. First, they might not systematically exist. Indeed, the majority of
population dynamics model fitting and risk forecasting assumes that population
dynamics are not heavy tailed (e.g.~REF, REF, REF). Alternatively, dynamics
might be heavy tailed, but a paucity of statistical methods for the frequently
available short time series has limited our ability to detect them. For
example, a commonly applied probability distribution to detect the shape of
time series tails (the generalized extreme value distribution) requires lumping
data into the most extreme event per unit time (e.g.~heaviest rainfall per
year) and therefore requires a considerable quantity of data (REF). Previous
large-scale work on the distribution of abundance trends focused on model
comparison to identify parsimonious descriptions, but has not quantified the
magnitude of heavy tails on a continuous scale (REF). Furthermore, information
criteria are poor at distinguishing subtle differences in population dynamics
for the relatively short time series that are most readily available (REF
CHANGE THIS?).
%bring in Ward - great method but data intensive, can't be applied to most
%individual time series from the GPDD
%plankton --- hideous distribution to fit, even harder in population dynamics
%framework -and what about seasonal cycles, density dependence?
%GPDD - model competition and finding most parsimonious explanation from set of
%hypotheses --- but only \emph{very} heavy from not heavy
%GEV --- super data intensive
%tipping point and regime shift methods --- say something\ldots{} not
%appropriate for most of the data available

Here we develop a probabilist framework for identifying ecological black swans
in population dynamic process noise --- the stochastic jumps from time step to
time step. Our framework quantifies both the magnitude and probability of
heavy-tailed dynamics, allows for a range of population dynamic models, can
incorporate observation uncertainty, and is applicable to commonly available
lengths of abundance time series. We apply our model to hundreds of populations
from around the world across four taxonomic classes to address three questions:
(1) how frequent and strong are black swans in population dynamics across
taxonomy, (2) what are some verified causes of black swans in ecological time
series, and (3) are there characteristics of time series or intrinsic
life-history characteristics associated with heavy-tailed dynamics. We find
a high probability of ecological black swans across a wide range of taxa but in
a limited number of populations. When they occur, extreme climate, predation,
parasites, and their interactions are common documented causes. We find little
evidence that intrinsic life-history characteristics are related to the
probability of black swans.
